Making Deep Q-learning Methods Robust to Time Discretization

### 2019-01-23 paper/icml_drau/framework.tex
.

%! TEX root = icml_drau.tex
\section{Near continuous-time reinforcement learning}
\label{sec:framework}

Many reinforcement learning environments are not intrinsically
time-discrete, but discretizations of an underlying continuous-time
environment. For instance, many simulated control environments, such as
the Mujoco environments~\cite{ddpg} or OpenAI Gym classic control
environments~\cite{gym}, are discretization of continuous-time control
problems.  In simulated environments, the time discretization is fixed by
the simulator, and is often used to approximate an underlying
differential equation.  In this case, the timestep may correspond to the
number of frames generated by second.  In real world environments,
sensors and actuators have a fixed time precision: cameras can only
capture a fixed amount of frames per second, and physical limitations
prevent actuators from responding instantaneously. The quality of these
components thus imposes a lower bound on the discretization timestep. As
the timestep $\deltat$ is not an intrinsic parameter of the
environment, but a constraint imposed by computational ressources, we
would expect that decreasing $\deltat$ could only improve the performance
of RL agents (though it might make optimization
harder).  RL algorithms should, at least, be resilient to a change of
$\deltat$, and should remain viable when $\deltat \rightarrow 0$.
Besides, designing a time discretization invariant algorithm could
alleviate tedious hyperparameterization by providing better defaults for
time-horizon-related parameters.

We are thus interested in the behavior of RL
algorithms in discretized environments, when the discretization timestep
is small. We will refer to such environments as \emph{near-continuous
environments}.
A formalized view of near-continuous environments is
given below, along with $\deltat$-dependent definitions of return, discount
factor and value functions, that converge to well defined
continous-time limits. The state-action value
function is shown to collapse to the value function as $\deltat$ goes to $0$.
Consequently there is no $Q$-learning in continuous time, foreshadowing
problematic behaviors of near continuous time $Q$-learning.
%      \TODO{We should explain somwhere here that, if not invariant,
%      harder to find good hyperparameters} DONE

@@

% TODO: For now, differential equation is the default, SDE is a footnote.
Let ${\cal S} = \mathbb{R}^d$ be a set of states, and ${\cal A}$ be a set of
actions. Consider the continuous \emph{Markov Decision Process} (MDP) defined by the differential equation %\TODO{Explain the intuition of Brownian in SDE}:
        \begin{equation}
	\label{eq:diffusion}
	ds_t/dt = F(s_t, a_t) % + \Sigma(s, a)\dbt.
              \end{equation}
where $F\colon {\cal S}\times{\cal A}\rightarrow {\cal S}$ describes
the dynamics of the environment. The agent interacts with the environment through a deterministic policy function
$\pi \colon {\cal S} \rightarrow {\cal A}$, so that $a_t=\pi(s_t)$. 
For simplicity we assume here that both the dynamics and exploitation policy are
deterministic; \footnote{
      We believe the results presented here hold more
      generally, assuming states
      follow a \emph{stochastic} differential equation \begin{equation}
	      ds = F(s, a)dt  + \Sigma(s, a)\dbt
	      \label{eq:sde}
      \end{equation} with
      $B_t$ a multidimensional Brownian motion and $\Sigma$ a covariance matrix. A
      formal treatment of SDEs is beyond the scope of this paper.}
 the exploration policy will be random, but care must be
taken to properly define random policies with
discrete actions in continuous time, see below.

For any $\deltat>0$, we can define an MDP ${\cal
M}_\deltat = \langle {\cal S}, {\cal A}, T_{\deltat}, r_\deltat,
      \gamma_\deltat\rangle$ as a discretization of the continuous MDP with
      \emph{time discretization} $\deltat$. The %deterministic
      transition function of a state $s$
      %$T_\deltat(s, a, s')$ 
      is %a dirac on $s_\deltat$, where $s_\deltat$ is 
      the
      state obtained
      when starting at $s_0 = s$ and maintaining $a_t=a$ constant for a time
      $\deltat$.
      This corresponds to an agent evolving in the continuous
      environment \eqref{eq:diffusion}, but 
      only making observations and choosing actions every $\deltat$. The
      rewards and decay factor are specified below. We
      call such an  MDP ${\cal M}_\deltat$
      \emph{near-continuous}.      % This framework covers classical control theory, as well as many usual simulated environments. TODO: not convinced that this is relevant here.

@@

By standard arguments for time discretizations of differential
equations, for a given policy the near-continuous states converge to the
continuous-time states when $\deltat\to 0$. \TODO{ref to appendix+formal
statement in appendix}% We now turn to the RL-related quantities.

@@

The definition~\eqref{eq:def-gamma} of the decay factor $\gamma_\deltat$ in
near-continuous environments keeps the time horizon constant in
\emph{physical} time, by making $\gamma_\deltat$ close to $1$ in
algorithmic time. Indeed, physical time horizon is $\deltat$ times the
algorithmic time horizon, namely
\begin{equation}
          \label{eq:time-horizon}
          \frac{\deltat}{1-\gamma^\deltat}
          %= - \frac{1}{\log \gamma} + \bigO(\deltat)
          \approx \frac{1}{1-\gamma},
        \end{equation}
which is thus stable when $\deltat\to 0$. 
On the other hand,
if $\gamma_\deltat$ was left constant
as $\deltat$ goes to $0$, the corresponding time horizon in physical time
would be $\approx \frac{\deltat}{1 - \gamma}$ which goes to $0$ when
$\deltat$ goes to $0$: such an agent would be increasingly short-sighted
as $\deltat\to 0$.

@@

Contrary to the value function, the action-value function $Q$ is ill-defined
for continuous-time MDPs. More precisely, the $Q$-function  collapses to the
$V$-function when $\deltat \rightarrow 0$. In near continuous time, the effect of
individual actions on the $Q$-function is of order $\bigO(\deltat)$. This
will make 
ranking of actions difficult, especially with an approximate
$Q$-function.
This argument appears informally in \cite{adv_upd}.
Formally:
\begin{theorem}
  \label{th:q-cont}
  Under suitable smoothness assumptions,
The action-value function of a near-continuous MDP is related to its
value function via
% Let $\pi$ be a deterministic policy such that for all $\deltat$, $V^\pi_\deltat$ is continuously
% differentiable, then, for all $s, a$,
\begin{equation}
	\label{eq:q-discr}
	Q^\pi_\deltat(s, a) = V^\pi_\deltat(s) + \bigO\left(\deltat\right)
\end{equation}
when $\deltat \to 0$ for every $(s,a)$.
\end{theorem}
% Consider a discretization of Eq.~\eqref{eq:cont_bell} with time interval
% $\deltat \ll~1$, Ito's lemma yield
% \begin{align}
% 	Q^\pi(s, a) &= \reward(s, a) \deltat + \gamma^{\deltat} \E_{\delta s \sim \delta s(s, a)}V^\pi(s + \delta s)\nonumber\\
% 		    &= \reward(s, a) \deltat + \gamma^{\deltat} V^\pi(s) +
% 		     \gamma^{\deltat} \partial_s V^\pi(s)f(s, a) \deltat\nonumber\\
% 		    &+ \frac{1}{2}\text{tr}\left(\Sigma^T\partial^2_{s^2}V^\pi(s)\Sigma\right)\deltat + \bigO(\deltat^2)\nonumber\\
% 		    &\xrightarrow[\deltat \to 0]{} V^\pi(s).
% 		    \label{eq:Q_cont}
                      %   \end{align}

>>>>

%! TEX root = icml_drau.tex
\section{Near Continuous-Time Reinforcement Learning}
\label{sec:framework}

Many reinforcement learning environments are not intrinsically
time-discrete, but discretizations of an underlying continuous-time
environment. For instance, many simulated control environments, such as
the Mujoco environments~\cite{ddpg} or OpenAI Gym classic control
environments~\cite{gym}, are discretizations of continuous-time control
problems.  In simulated environments, the time discretization is fixed by
the simulator, and is often used to approximate an underlying
differential equation.  In this case, the timestep may correspond to the
number of frames generated by second.  In real world environments,
sensors and actuators have a fixed time precision: cameras can only
capture a fixed amount of frames per second, and physical limitations
prevent actuators from responding instantaneously. The quality of these
components thus imposes a lower bound on the discretization timestep. As
the timestep $\deltat$ is 
%not an intrinsic parameter of the environment, but
largely a constraint imposed by computational ressources, we
would expect that decreasing $\deltat$ would only improve the performance
of RL agents (though it might make optimization
harder).  RL algorithms should, at least, be resilient to a change of
$\deltat$, and should remain viable when $\deltat \rightarrow 0$.
Besides, designing a time discretization invariant algorithm could
alleviate tedious hyperparameterization by providing better defaults for
time-horizon-related parameters.

We are thus interested in the behavior of RL
algorithms in discretized environments, when the discretization timestep
is small. We will refer to such environments as \emph{near-continuous
environments}.
A formalized view of near-continuous environments is
given below, along with $\deltat$-dependent definitions of return, discount
factor and value functions, that converge to well defined
continous-time limits. The state-action value
function is shown to collapse to the value function as $\deltat$ goes to $0$.
Consequently there is no $Q$-learning in continuous time, foreshadowing
problematic behavior of $Q$-learning with small timesteps.
%      \TODO{We should explain somwhere here that, if not invariant,
%      harder to find good hyperparameters} DONE

@@

% TODO: For now, differential equation is the default, SDE is a footnote.
Let ${\cal S} = \mathbb{R}^d$ be a set of states, and ${\cal A}$ be a set of
actions. Consider the continuous-time \emph{Markov Decision Process} (MDP) defined by the differential equation %\TODO{Explain the intuition of Brownian in SDE}:
        \begin{equation}
	\label{eq:diffusion}
	ds_t/dt = F(s_t, a_t) % + \Sigma(s, a)\dbt.
              \end{equation}
where $F\colon {\cal S}\times{\cal A}\rightarrow {\cal S}$ describes
the dynamics of the environment. The agent interacts with the environment through a deterministic policy function
$\pi \colon {\cal S} \rightarrow {\cal A}$, so that $a_t=\pi(s_t)$;
actions can be discrete or continuous. 
For simplicity we assume here that both the dynamics and exploitation policy are
deterministic; \footnote{
      We believe the results presented here hold more
      generally, assuming states
      follow a \emph{stochastic} differential equation \begin{equation}
	      ds = F(s, a)dt  + \Sigma(s, a)\dbt
	      \label{eq:sde}
      \end{equation} with
      $B_t$ a multidimensional Brownian motion and $\Sigma$ a covariance matrix. A
      formal treatment of SDEs is beyond the scope of this paper.}
 the exploration policy will be random, but care must be
taken to define proper random policies 
in continuous time, especially with discrete actions (Sec.~\ref{subsec:explo}).

For any timestep $\deltat>0$, we can define an MDP ${\cal
M}_\deltat = \langle {\cal S}, {\cal A}, T_{\deltat}, r_\deltat,
      \gamma_\deltat\rangle$ as a discretization of the continuous-time MDP with
      \emph{time discretization} $\deltat$. The %deterministic
      transition function of a state $s$
      %$T_\deltat(s, a, s')$ 
      is %a dirac on $s_\deltat$, where $s_\deltat$ is 
      the
      state obtained
      when starting at $s_0 = s$ and maintaining $a_t=a$ constant for a time
      $\deltat$.
      This corresponds to an agent evolving in the continuous
      environment \eqref{eq:diffusion}, but 
      only making observations and choosing actions every $\deltat$. The
      rewards and decay factor are specified below. We
      call such an  MDP ${\cal M}_\deltat$
      \emph{near-continuous}.      % This framework covers classical control theory, as well as many usual simulated environments. TODO: not convinced that this is relevant here.

@@

% By standard arguments for time discretizations of differential
% equations, for a given policy the near-continuous states converge to the
% continuous-time states when $\deltat\to 0$. \TODO{ref to appendix+formal
% statement in appendix}% We now turn to the RL-related quantities.

@@

The definition~\eqref{eq:def-gamma} of the decay factor $\gamma_\deltat$ in
near-continuous environments keeps the time horizon constant in
\emph{physical} time, by making $\gamma_\deltat$ close to $1$ in
algorithmic time. Indeed, physical time horizon is $\deltat$ times the
algorithmic time horizon, namely
\begin{equation}
          \label{eq:time-horizon}
          \frac{\deltat}{1-\gamma^\deltat}
          = - \frac{1}{\log \gamma} + \bigO(\deltat)
          \approx \frac{1}{1-\gamma},
        \end{equation}
which is thus stable when $\deltat\to 0$. 
On the other hand,
if $\gamma_\deltat$ was left constant
as $\deltat$ goes to $0$, the corresponding time horizon in physical time
would be $\approx \frac{\deltat}{1 - \gamma}$ which goes to $0$ when
$\deltat$ goes to $0$: such an agent would be increasingly short-sighted
as $\deltat\to 0$.

@@

Contrary to the value function, the action-value function $Q$ is ill-defined
for continuous-time MDPs. More precisely, the $Q$-function  collapses to the
$V$-function when $\deltat \rightarrow 0$. In near continuous time, the effect of
individual actions on the $Q$-function is of order $\bigO(\deltat)$. This
will make 
ranking of actions difficult, especially with an approximate
$Q$-function.
This argument appears informally in \cite{adv_upd}.
Formally:
\begin{theorem}
  \label{th:q-cont}
  Under suitable smoothness assumptions,
The action-value function of a near-continuous MDP is related to its
value function via
% Let $\pi$ be a deterministic policy such that for all $\deltat$, $V^\pi_\deltat$ is continuously
% differentiable, then, for all $s, a$,
\begin{equation}
	\label{eq:q-discr}
	Q^\pi_\deltat(s, a) = V^\pi_\deltat(s) + \bigO\left(\deltat\right)
\end{equation}
when $\deltat \to 0$, for every $(s,a)$.
\end{theorem}
% Consider a discretization of Eq.~\eqref{eq:cont_bell} with time interval
% $\deltat \ll~1$, Ito's lemma yield
% \begin{align}
% 	Q^\pi(s, a) &= \reward(s, a) \deltat + \gamma^{\deltat} \E_{\delta s \sim \delta s(s, a)}V^\pi(s + \delta s)\nonumber\\
% 		    &= \reward(s, a) \deltat + \gamma^{\deltat} V^\pi(s) +
% 		     \gamma^{\deltat} \partial_s V^\pi(s)f(s, a) \deltat\nonumber\\
% 		    &+ \frac{1}{2}\text{tr}\left(\Sigma^T\partial^2_{s^2}V^\pi(s)\Sigma\right)\deltat + \bigO(\deltat^2)\nonumber\\
% 		    &\xrightarrow[\deltat \to 0]{} V^\pi(s).
% 		    \label{eq:Q_cont}
                      %   \end{align}

### 2019-01-23 paper/icml_drau/intro.tex

\section{Introduction}
\label{sec:intro}
In recent years, \emph{Deep Reinforcement Learning} (DRL) approaches have
provided impressive results in a variety of domains, achieving superhuman
performance with no expert knowledge in perfect information zero-sum
games~\cite{alphazero}, reaching level of top players in video
games~(\citealt{openai_five}, \citealt{dqn}), or learning dexterous manipulation
from scratch without demonstrations~\cite{hand_control}. 
In spite of those successes, DRL approaches are
sensitive to a number of factors, including hyperparameterization,
implementation details or small changes in the environment
parameters~(\citealt{drl_matter}, \citealt{drl_matter_bis}). This sensitivity,
along with sample inefficiency, largely prevents DRL from being applied in real
world settings. Notably, high sensitivity to environment parameters prevents
transfer from imperfect simulators to real world scenarios.

@@

Our contribution is threefold:
\begin{itemize} 
\item Building on \cite{adv_upd}, we formally show that the $Q$-function collapses to the $V$-function in near-continuous time, and thus that
    standard $Q$-learning is ill-behaved in this setting.
  \item Our analysis of limit properties in near continuous environments leads to a robust off-policy algorithm. In particular,
    it provides insights on how to design architectures, constrain exploration, and scale learning rates.
  \item We empirically show that standard $Q$-learning methods are not
  robust to changes in time discretization, exhibiting degraded
  performance, while our algorithm demonstrates
  substantial robustness.
\end{itemize}

>>>>

\section{Introduction}
\label{sec:intro}
In recent years, \emph{Deep Reinforcement Learning} (DRL) approaches have
provided impressive results in a variety of domains, achieving superhuman
performance with no expert knowledge in perfect information zero-sum
games~\cite{alphazero}, reaching top player level in video
games~(\citealt{openai_five}, \citealt{dqn}), or learning dexterous manipulation
from scratch without demonstrations~\cite{hand_control}. 
In spite of those successes, DRL approaches are
sensitive to a number of factors, including hyperparameterization,
implementation details or small changes in the environment
parameters~(\citealt{drl_matter}, \citealt{drl_matter_bis}). This sensitivity,
along with sample inefficiency, largely prevents DRL from being applied in real
world settings. Notably, high sensitivity to environment parameters prevents
transfer from imperfect simulators to real world scenarios.

@@

Our contribution is threefold:
\begin{itemize} 
\item Building on \cite{adv_upd}, we formally show that the $Q$-function collapses to the $V$-function in near-continuous time, and thus that
    standard $Q$-learning is ill-behaved in this setting.
  \item Our analysis of properties in the continuous-time limit leads to a robust off-policy algorithm. In particular,
    it provides insights on architecture design, and constrains
    exploration schemes and learning rates scalings.
  \item We empirically show that standard $Q$-learning methods are not
  robust to changes in time discretization, exhibiting degraded
  performance, while our algorithm demonstrates
  substantial robustness.
\end{itemize}

### 2019-01-23 paper/icml_drau/relatedworks.tex

Our approach builds on \cite{adv_upd},  who identified the collapse of
$Q$-learning for small time steps and, as a solution, suggested the Advantage
Updating algorithm, with proper $\deltat$ scalings for the $V$ part and
the advantage part; testing was only done on a quadratic-linear problem.
%\TODO{mention that Baird did not experimentally study robustness. NDY:
%fifth point below}

We expand on \cite{adv_upd} in several directions. First, we modify the
algorithm by
using a different normalization step for $A$, which forgoes the need to
learn the normalization itself, thanks to the parameterization
\eqref{eq:Aparam}. Second, we test Advantage Updating for the first time on a
variety on RL environments using deep networks, establishing Deep
Advantage Updating as a viable algorithm in this setting.  Third,
we provide formal proofs in a general setting for the collapse of $Q$-learning when $\deltat\to
0$, and for the non-collapse of Advantage Updating with the proper scalings.
Fourth, we also discuss how to obtain $\deltat$-invariant exploration. Fifth, we
provide stringent experimental tests of the actual robustness to changing
$\deltat$.
%  \cite{adv_upd} already separated the learning procedure of the
% value and advantage functions, and introduced proper scalings between the two
% components. We extend on \cite{adv_upd} in three directions. We provide a
% formal argument on why the advantage contribution of the $Q$-function is of order
% $\bigO(\deltat)$, irrelevant of the stochasticity of the environment, in a quite
% general framework. We show that, provided with well suited learning rates and
% exploration method, Advantage updating provide near invariance to change of time
% discretization. We show that Advantage updating is viable in the context of deep
% reinforcement learning, while not using the true gradient of the residual error.

Our study focuses on off-policy algorithms. Some on-policy algorithms, such as
A3C~\cite{a3c}, or PPO~\cite{ppo} may be time discretization invariant in
specific setups. This is out of the scope of our work and would require a
separate study.

>>>>

Our approach builds on \cite{adv_upd},  who identified the collapse of
$Q$-learning for small time steps and, as a solution, suggested the Advantage
Updating algorithm, with proper scalings for the $V$ and
advantage parts depending on timescale $\deltat$; testing was only done on a quadratic-linear problem.
%\TODO{mention that Baird did not experimentally study robustness. NDY:
%fifth point below}

We expand on \cite{adv_upd} in several directions. First, we modify the
algorithm by
using a different normalization step for $A$, which forgoes the need to
learn the normalization itself, thanks to the parameterization
\eqref{eq:Aparam}. Second, we test Advantage Updating for the first time on a
variety on RL environments using deep networks, establishing Deep
Advantage Updating as a viable algorithm in this setting.  Third,
we provide formal proofs in a general setting for the collapse of
$Q$-learning when the timescale $\deltat$ tends to $0$,
and for the non-collapse of Advantage Updating with the proper scalings.
Fourth, we also discuss how to obtain $\deltat$-invariant exploration. Fifth, we
provide stringent experimental tests of the actual robustness to changing
$\deltat$.
%  \cite{adv_upd} already separated the learning procedure of the
% value and advantage functions, and introduced proper scalings between the two
% components. We extend on \cite{adv_upd} in three directions. We provide a
% formal argument on why the advantage contribution of the $Q$-function is of order
% $\bigO(\deltat)$, irrelevant of the stochasticity of the environment, in a quite
% general framework. We show that, provided with well suited learning rates and
% exploration method, Advantage updating provide near invariance to change of time
% discretization. We show that Advantage updating is viable in the context of deep
% reinforcement learning, while not using the true gradient of the residual error.

Our study focuses on off-policy algorithms. Some on-policy algorithms,
such as A3C~\cite{a3c}, or PPO~\cite{ppo} may be time discretization
invariant with specific setups. This is out of the scope of our work and
would require a separate study.
